---
title: "Groundfish Survey Design DA"
author: "Jessica Zeng"
date: "2026-08-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("readxl")
install.packages("corrplot")
```

```{r}
library(readxl)
library(corrplot)
```

```{r}
# install.packages("doBy")
# install.packages("pbkrtest")
# install.packages("car")
# 
# library(car)
```

## Data Analysis

```{r}
# Read in data
df <- read_excel("spring_historical_english.xlsx")
```

```{r}
str(df)
summary(df)
```
Look at how many unique variables there are
```{r}
sapply(df, function(x) length(unique(x)))
```

Remove useless prediction variables
```{r}
df_model_general <- df[, !names(df) %in% c(
  "mission_id",
  "mission_name",
  "station_id",
  "station_type",
  "station_number",
  "date",
  "cast_latitude_DMS",
  "cast_longitude_DMS",
  "haul_latitude_DMS",
  "haul_longitude_DMS",
  "sample_id",
  "londunarhofn_nr",
  "name",
  "month",
  "region",
  "sub_region",
  "gear_number",
  "gear_name",
  "sample_category",
  "trawl_number",
  "trawl_start",
  "trawl_end",
  "fishing_grounds",
  "einkst",
  "einknr",
  "flokkur",
  "heimah",
  "current_dir",
  "current_speed",
  "ocean_state",
  "brl"
)]
```

NOTE: we will remove tow_station in the future, as it is less of a predictor variable, and more of a grouping variable. We will factor this in later.

```{r}
str(df_model_general)
```
Remaining variables:
year
cast_latitude
cast_longitude
haul_latitude
haul_longitude
depth_cast
depth_haul
ship_number
wind_speed
wind_direction
weather
cloud
sea_state
bottom_temp
air_temp
barometer
trawl_speed
trawl_length
trawl_bearing
trawl_time
catch
tow_station
quadrant

## Exploration
### Numerical variables
```{r}
plot_catch_relationship <- function(variable, data) {
  
  plot(data[[variable]], data$catch,
       main = paste("Catch vs", variable),
       xlab = variable,
       ylab = "Catch (kg)",
       pch = 19)
  
  abline(lm(data$catch ~ data[[variable]]),
         lwd = 2)
}
```

```{r}
plot_catch_relationship("year", df_model_general)
plot_catch_relationship("depth_cast", df_model_general)
plot_catch_relationship("depth_haul", df_model_general)
plot_catch_relationship("cast_latitude", df_model_general)
plot_catch_relationship("cast_longitude", df_model_general)
plot_catch_relationship("haul_latitude", df_model_general)
plot_catch_relationship("haul_longitude", df_model_general)
plot_catch_relationship("bottom_temp", df_model_general)
plot_catch_relationship("surface_temp", df_model_general)
plot_catch_relationship("air_temp", df_model_general)
plot_catch_relationship("barometer", df_model_general)
plot_catch_relationship("wind_speed", df_model_general)
plot_catch_relationship("trawl_speed", df_model_general)
plot_catch_relationship("trawl_length", df_model_general)
plot_catch_relationship("trawl_time", df_model_general)
plot_catch_relationship("trawl_bearing", df_model_general)
```

One large plot for the report:
```{r}
plot_catch_relationship <- function(variable, data) {
  
  plot(data[[variable]], data$catch,
       main = paste("Catch vs", variable),
       xlab = variable,
       ylab = "Catch (kg)",
       pch = 19,
       cex = 0.5)
  
  abline(lm(data$catch ~ data[[variable]]),
         lwd = 2)
}

# Set up a 4 x 4 grid
par(mfrow = c(4, 4),
    mar = c(4, 4, 3, 1))

plot_catch_relationship("year", df_model_general)
plot_catch_relationship("depth_cast", df_model_general)
plot_catch_relationship("depth_haul", df_model_general)
plot_catch_relationship("cast_latitude", df_model_general)
plot_catch_relationship("cast_longitude", df_model_general)
plot_catch_relationship("haul_latitude", df_model_general)
plot_catch_relationship("haul_longitude", df_model_general)
plot_catch_relationship("bottom_temp", df_model_general)
plot_catch_relationship("surface_temp", df_model_general)
plot_catch_relationship("air_temp", df_model_general)
plot_catch_relationship("barometer", df_model_general)
plot_catch_relationship("wind_speed", df_model_general)
plot_catch_relationship("trawl_speed", df_model_general)
plot_catch_relationship("trawl_length", df_model_general)
plot_catch_relationship("trawl_time", df_model_general)
plot_catch_relationship("trawl_bearing", df_model_general)

# Reset plotting layout
par(mfrow = c(1, 1))
```


### Categorical variables
```{r}
# Convert categorical variables to factors
df_model_general$ship_number <- as.factor(df_model_general$ship_number)
df_model_general$quadrant <- as.factor(df_model_general$quadrant)
df_model_general$weather <- as.factor(df_model_general$weather)
df_model_general$cloud <- as.factor(df_model_general$cloud)
df_model_general$sea_state <- as.factor(df_model_general$sea_state)
```

```{r}
# Catch by quadrant
boxplot(catch ~ quadrant,
        data = df_model_general,
        main = "Catch by Quadrant",
        xlab = "Quadrant",
        ylab = "Catch (kg)")
# Catch by ship
boxplot(catch ~ ship_number,
        data = df_model_general,
        main = "Catch by Ship",
        xlab = "Ship Number",
        ylab = "Catch (kg)")
# Catch by weather condition
boxplot(catch ~ weather,
        data = df_model_general,
        main = "Catch by Weather",
        xlab = "Weather",
        ylab = "Catch (kg)")
# Catch by cloud condition
boxplot(catch ~ cloud,
        data = df_model_general,
        main = "Catch by Cloud Condition",
        xlab = "Cloud",
        ylab = "Catch (kg)")
# Catch by sea state
boxplot(catch ~ sea_state,
        data = df_model_general,
        main = "Catch by Sea State",
        xlab = "Sea State",
        ylab = "Catch (kg)")
```
One plot for report:
```{r}
# Set up a 2 x 3 grid
par(mfrow = c(2, 3),
    mar = c(4, 4, 3, 1))

# Catch by quadrant
boxplot(catch ~ quadrant,
        data = df_model_general,
        main = "Catch by Quadrant",
        xlab = "Quadrant",
        ylab = "Catch (kg)")

# Catch by ship
boxplot(catch ~ ship_number,
        data = df_model_general,
        main = "Catch by Ship",
        xlab = "Ship Number",
        ylab = "Catch (kg)")

# Catch by weather condition
boxplot(catch ~ weather,
        data = df_model_general,
        main = "Catch by Weather",
        xlab = "Weather",
        ylab = "Catch (kg)")

# Catch by cloud condition
boxplot(catch ~ cloud,
        data = df_model_general,
        main = "Catch by Cloud Condition",
        xlab = "Cloud",
        ylab = "Catch (kg)")

# Catch by sea state
boxplot(catch ~ sea_state,
        data = df_model_general,
        main = "Catch by Sea State",
        xlab = "Sea State",
        ylab = "Catch (kg)")

# Reset plotting layout
par(mfrow = c(1, 1))
```


## Preliminary model
```{r}
# Variables for preliminary model
df_prelim <- df_model_general[, c(
  "year",
  "cast_latitude",
  "cast_longitude",
  "haul_latitude",
  "haul_longitude",
  "depth_cast",
  "depth_haul",
  "ship_number",
  "wind_speed",
  "wind_direction",
  "weather",
  "cloud",
  "sea_state",
  "bottom_temp",
  "air_temp",
  "barometer",
  "trawl_speed",
  "trawl_length",
  "trawl_bearing",
  "trawl_time",
  "catch",
  "tow_station",
  "quadrant"
)]
```

Treat as categorical variables
```{r}
df_prelim$ship_number <- as.factor(df_prelim$ship_number)
df_prelim$weather <- as.factor(df_prelim$weather)
df_prelim$cloud <- as.factor(df_prelim$cloud)
df_prelim$sea_state <- as.factor(df_prelim$sea_state)
df_prelim$quadrant <- as.factor(df_prelim$quadrant)
```

### Examine catch
```{r}
# Raw catch
hist(df_prelim$catch,
     breaks = 50,
     main = "Distribution of Raw Catch",
     xlab = "Catch (kg)")

# Log-transformed catch
df_prelim$log_catch <- log(df_prelim$catch)

hist(df_prelim$log_catch,
     breaks = 50,
     main = "Distribution of Log(Catch)",
     xlab = "Log(Catch) (kg)")
```

### Examine trawl variables
```{r}
#Correlation matrix
trawl_cor <- cor(
  df_prelim[, c(
    "trawl_speed",
    "trawl_length",
    "trawl_time"
  )],
  use = "pairwise.complete.obs"
)

round(trawl_cor, 3)
```
There is essentially no linear correlation between trawl_speed and trawl_length (r=0.055)

There is a moderate negative relationship between trawl_speed and trawl_time (r=-0.483)

There is a strong positive relationship between trawl_length and trawl_time (r=0.768)


Look into relationship between length & time
```{r}
plot(
  df_prelim$trawl_length,
  df_prelim$trawl_time,
  main = "Trawl Length vs Trawl Time",
  xlab = "Trawl Length",
  ylab = "Trawl Time",
  pch = 19
)

abline(
  lm(trawl_time ~ trawl_length,
     data = df_prelim),
  lwd = 2
)
```


### Look into correlations between numerical variables
```{r}
numeric_vars <- df_prelim[, c(
  "year",
  "cast_latitude",
  "cast_longitude",
  "haul_latitude",
  "haul_longitude",
  "depth_cast",
  "depth_haul",
  "wind_speed",
  "wind_direction",
  "bottom_temp",
  "air_temp",
  "barometer",
  "trawl_speed",
  "trawl_length",
  "trawl_bearing",
  "trawl_time"
)]

cor_matrix <- cor(
  numeric_vars,
  use = "pairwise.complete.obs"
)

round(cor_matrix, 2)

corrplot(
  cor_matrix,
  method = "circle",
  type = "upper",
  tl.col = "black",
  tl.srt = 45
)
```

```{r}
preliminary_model <- lm(
  log_catch ~
    year +
    cast_latitude +
    cast_longitude +
    haul_latitude +
    haul_longitude +
    depth_cast +
    depth_haul +
    ship_number +
    wind_speed +
    wind_direction +
    weather +
    cloud +
    sea_state +
    bottom_temp +
    air_temp +
    barometer +
    trawl_speed +
    trawl_length +
    trawl_bearing +
    trawl_time +
    quadrant,
  data = df_prelim
)

summary(preliminary_model)
```

### Check multi collinearity (VIF)
car library not working, make own vif function:
```{r}
# Calculate VIF for each numerical predictor
calculate_vif <- function(model) {
  
  # Get the model matrix
  X <- model.matrix(model)[, -1]
  
  # Calculate VIF for each column
  vif_values <- numeric(ncol(X))
  names(vif_values) <- colnames(X)
  
  for (i in seq_along(vif_values)) {
    
    other_vars <- X[, -i, drop = FALSE]
    target <- X[, i]
    
    auxiliary_model <- lm(target ~ other_vars)
    
    R2 <- summary(auxiliary_model)$r.squared
    
    vif_values[i] <- 1 / (1 - R2)
  }
  
  return(vif_values)
}
```

```{r}
vif_values <- calculate_vif(preliminary_model)
round(vif_values, 2)
```

Rough interpretation of VIF
VIF 1-2, little concern
VIF 2-5, moderate
VIF 5-10, high
VIF >10 Very high

Can remove cast_latitude, cast_longitude as they are highly correlated with haul_latitude, haul_longitude.

Can remove depth_cast as similar to depth_haul 

Even though fround before that trawl speed, time and length are their own variables, the VIF finds speed and time redundant relative to the other trawl variables. Therefore we will remove trawl_speed and trawl_time

Remove quadrant too as it was created from the latitude and longitude variables. This variable may be useful for more simpler models, but not for this multilinear regression model. 

```{r}
# df_prelim$cast_latitude <- NULL
# df_prelim$cast_latitude <- NULL
# df_prelim$depth_cast <- NULL
```

For sake of simplification, we will remove weather, cloud, sea_state, wind_speed, and wind_direction. However, with more time, would try to investigate these variables further. 


## Build second model

### Build new dataset
```{r}
df_model_2 <- df_prelim[, c(
  "year",
  "haul_latitude",
  "haul_longitude",
  "depth_haul",
  "ship_number",
  "bottom_temp",
  "air_temp",
  "barometer",
  "trawl_length",
  "trawl_bearing",
  "catch",
  "log_catch",
  "tow_station"
)]
```

### Build new model
```{r}
model_2 <- lm(
  log_catch ~
    year +
    haul_latitude +
    haul_longitude +
    depth_haul +
    ship_number +
    bottom_temp +
    air_temp +
    barometer +
    trawl_length +
    trawl_bearing,
  data = df_model_2
)

summary(model_2)
```
Model 2 explains 18.36% of variation, but model 1 explains 20.9%.


### Check collinearity
```{r}
vif_values_2 <- calculate_vif(model_2)

round(vif_values_2, 2)
```
Largest VIF is 4.02, which is generally acceptable. 

We can now use model 2 as a baseline model. 

```{r}
summary(model_2)$coefficients
```
barometer is not significant with p=0.897
Similarly, depth_haul also has a high p-value of p=0.964. However this may suggest that the variable is non-linearly correlated with catch.

```{r}
par(mfrow = c(2, 2))
plot(model_2)
```

```{r}
drop1(model_2, test = "F")
```

## TESTING
Test model 2 and model 1
1. Extract 2002-2022, divide into 10 years training/testing
2. Fit regression model
3. Predict catch for testing years
4. Generate 1000 surveys
5. Repeat 100 times

```{r}
library(ggplot2)
```

Set up both models
```{r}
library(dplyr)

df <- read_excel("spring_historical_english.xlsx")

# Create log catch
df <- df |>
  mutate(
    catch = as.numeric(catch),
    log_catch = log(catch)
  )

# Formulas from your two fitted models
model_formulas <- list(
  "Preliminary Model" = formula(preliminary_model),
  "Model 2" = formula(model_2)
)

# All predictors required by both models
required_variables <- unique(
  unlist(lapply(model_formulas, all.vars))
)

# Keep rows with valid response and predictor values
df_test <- df |>
  filter(
    catch > 0,
    is.finite(catch),
    is.finite(log_catch),
    if_all(all_of(required_variables), ~ !is.na(.x))
  )

# Restrict testing to the chosen historical period
years <- sort(unique(
  df_test$year[
    df_test$year >= 2002 &
    df_test$year <= 2022
  ]
))

years <- tail(years, 20)

if (length(years) != 20) {
  stop("There must be 20 usable years in the selected period.")
}

df_test <- df_test |>
  filter(year %in% years)
```


Run both models on same tests
```{r}
set.seed(566)

n_iterations <- 100
n_simulations <- 1000

results_list <- vector("list", n_iterations * 2)
result_index <- 1

for (i in seq_len(n_iterations)) {

  # ----------------------------------------
  # Step 1: Shared training/testing split
  # ----------------------------------------

  training_years <- sample(years, 10)
  testing_years <- setdiff(years, training_years)

  train_data <- df_test |>
    filter(year %in% training_years)

  test_data <- df_test |>
    filter(year %in% testing_years)

  # Actual mean annual total catch
  actual_annual_totals <- test_data |>
    group_by(year) |>
    summarise(
      total_catch = sum(catch),
      .groups = "drop"
    )

  actual_mean <- mean(actual_annual_totals$total_catch)

  # ----------------------------------------
  # Step 2: Fit and evaluate each model
  # ----------------------------------------

  for (model_name in names(model_formulas)) {

    model_formula <- model_formulas[[model_name]]

    # Fit model using training years only
    fitted_model <- lm(
      model_formula,
      data = train_data
    )

    # Predict log catch for the testing years
    predicted_log_catch <- predict(
      fitted_model,
      newdata = test_data
    )

    # Training residuals only
    training_residuals <- residuals(fitted_model)

    # --------------------------------------
    # Step 3: Simulate survey catches
    # --------------------------------------

    simulated_mean_totals <- numeric(n_simulations)

    for (j in seq_len(n_simulations)) {

      # Resample residuals on the log scale
      simulated_residuals <- sample(
        training_residuals,
        size = nrow(test_data),
        replace = TRUE
      )

      simulated_catch <- exp(
        predicted_log_catch + simulated_residuals
      )

      simulated_data <- data.frame(
        year = test_data$year,
        catch = simulated_catch
      )

      # Total catch for each testing year
      simulated_annual_totals <- simulated_data |>
        group_by(year) |>
        summarise(
          total_catch = sum(catch),
          .groups = "drop"
        )

      # Mean annual survey total
      simulated_mean_totals[j] <- mean(
        simulated_annual_totals$total_catch
      )
    }

    # --------------------------------------
    # Step 4: Calculate simulation interval
    # --------------------------------------

    lower_95 <- unname(quantile(
      simulated_mean_totals, 0.025
    ))

    upper_95 <- unname(quantile(
      simulated_mean_totals, 0.975
    ))

    predicted_mean <- mean(simulated_mean_totals)

    # --------------------------------------
    # Step 5: Evaluate performance
    # --------------------------------------

    results_list[[result_index]] <- data.frame(
      iteration = i,
      model = model_name,
      predicted_mean = predicted_mean,
      actual_mean = actual_mean,
      lower_95 = lower_95,
      upper_95 = upper_95,
      covered = actual_mean >= lower_95 &&
                actual_mean <= upper_95,
      absolute_percentage_error =
        abs(predicted_mean - actual_mean) /
        actual_mean * 100,
      interval_width = upper_95 - lower_95
    )

    result_index <- result_index + 1
  }

  if (i %% 10 == 0) {
    cat("Completed iteration", i, "of", n_iterations, "\n")
  }
}

# Combine results from both models
results <- bind_rows(results_list)

head(results)
```

```{r}

# Combine results from both models
results <- bind_rows(results_list)

head(results)
```

Compare models
```{r}
model_comparison <- results |>
  group_by(model) |>
  summarise(
    iterations = n(),

    coverage_percent = mean(covered) * 100,

    mean_absolute_percentage_error =
      mean(absolute_percentage_error),

    median_absolute_percentage_error =
      median(absolute_percentage_error),

    mean_interval_width =
      mean(interval_width),

    mean_relative_interval_width_percent =
      mean(interval_width / actual_mean * 100),

    .groups = "drop"
  )

print(model_comparison)
```


Plot models
```{r}
ggplot(results, aes(x = iteration)) +

  geom_linerange(
    aes(ymin = lower_95, ymax = upper_95),
    colour = "steelblue",
    linewidth = 0.5
  ) +

  geom_point(
    aes(y = predicted_mean),
    colour = "steelblue",
    size = 1.3
  ) +

  geom_point(
    aes(y = actual_mean),
    colour = "red",
    size = 1.3
  ) +

  facet_wrap(
    ~ model,
    ncol = 1,
    labeller = as_labeller(c(
      "Preliminary Model" = "Preliminary Model",
      "Model 2" = "Simplified Model"
    ))
  ) +

  labs(
    title = "Regression Models: Simulated Mean Catch and 95% Intervals",
    subtitle = "Blue = simulated mean; red = observed mean",
    x = "Testing iteration",
    y = "Mean annual total catch (kg)"
  ) +

  theme_minimal() +

  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```


