---
title: "Groundfish Survey Design DA"
author: "Jessica Zeng"
date: "2026-08-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r}
install.packages("readxl")
library(readxl)

install.packages("corrplot")
library(corrplot)
```
```{r}
# install.packages("doBy")
# install.packages("pbkrtest")
# install.packages("car")
# 
# library(car)
```

## Data Analysis

```{r}
# Read in data
df <- read_excel("spring_historical_english.xlsx")
```

```{r}
str(df)
summary(df)
```
Look at how many unique variables there are
```{r}
sapply(df, function(x) length(unique(x)))
```

Remove useless prediction variables
```{r}
df_model_general <- df[, !names(df) %in% c(
  "mission_id",
  "mission_name",
  "station_id",
  "station_type",
  "station_number",
  "date",
  "cast_latitude_DMS",
  "cast_longitude_DMS",
  "haul_latitude_DMS",
  "haul_longitude_DMS",
  "sample_id",
  "londunarhofn_nr",
  "name",
  "month",
  "region",
  "sub_region",
  "gear_number",
  "gear_name",
  "sample_category",
  "trawl_number",
  "trawl_start",
  "trawl_end",
  "fishing_grounds",
  "einkst",
  "einknr",
  "flokkur",
  "heimah",
  "current_dir",
  "current_speed",
  "ocean_state",
  "brl"
)]
```

NOTE: we will remove tow_station in the future, as it is less of a predictor variable, and more of a grouping variable. We will factor this in later.

```{r}
str(df_model_general)
```
Remaining variables:
year
cast_latitude
cast_longitude
haul_latitude
haul_longitude
depth_cast
depth_haul
ship_number
wind_speed
wind_direction
weather
cloud
sea_state
bottom_temp
air_temp
barometer
trawl_speed
trawl_length
trawl_bearing
trawl_time
catch
tow_station
quadrant

## Exploration
### Numerical variables
```{r}
plot_catch_relationship <- function(variable, data) {
  
  plot(data[[variable]], data$catch,
       main = paste("Catch vs", variable),
       xlab = variable,
       ylab = "Catch (kg)",
       pch = 19)
  
  abline(lm(data$catch ~ data[[variable]]),
         lwd = 2)
}
```

```{r}
plot_catch_relationship("year", df_model_general)
plot_catch_relationship("depth_cast", df_model_general)
plot_catch_relationship("depth_haul", df_model_general)
plot_catch_relationship("cast_latitude", df_model_general)
plot_catch_relationship("cast_longitude", df_model_general)
plot_catch_relationship("haul_latitude", df_model_general)
plot_catch_relationship("haul_longitude", df_model_general)
plot_catch_relationship("bottom_temp", df_model_general)
plot_catch_relationship("surface_temp", df_model_general)
plot_catch_relationship("air_temp", df_model_general)
plot_catch_relationship("barometer", df_model_general)
plot_catch_relationship("wind_speed", df_model_general)
plot_catch_relationship("trawl_speed", df_model_general)
plot_catch_relationship("trawl_length", df_model_general)
plot_catch_relationship("trawl_time", df_model_general)
plot_catch_relationship("trawl_bearing", df_model_general)
```

### Categorical variables
```{r}
# Convert categorical variables to factors
df_model_general$ship_number <- as.factor(df_model_general$ship_number)
df_model_general$quadrant <- as.factor(df_model_general$quadrant)
df_model_general$weather <- as.factor(df_model_general$weather)
df_model_general$cloud <- as.factor(df_model_general$cloud)
df_model_general$sea_state <- as.factor(df_model_general$sea_state)
```

```{r}
# Catch by quadrant
boxplot(catch ~ quadrant,
        data = df_model_general,
        main = "Catch by Quadrant",
        xlab = "Quadrant",
        ylab = "Catch (kg)")
# Catch by ship
boxplot(catch ~ ship_number,
        data = df_model_general,
        main = "Catch by Ship",
        xlab = "Ship Number",
        ylab = "Catch (kg)")
# Catch by weather condition
boxplot(catch ~ weather,
        data = df_model_general,
        main = "Catch by Weather",
        xlab = "Weather",
        ylab = "Catch (kg)")
# Catch by cloud condition
boxplot(catch ~ cloud,
        data = df_model_general,
        main = "Catch by Cloud Condition",
        xlab = "Cloud",
        ylab = "Catch (kg)")
# Catch by sea state
boxplot(catch ~ sea_state,
        data = df_model_general,
        main = "Catch by Sea State",
        xlab = "Sea State",
        ylab = "Catch (kg)")
```


## Preliminary model
```{r}
# Variables for preliminary model
df_prelim <- df_model_general[, c(
  "year",
  "cast_latitude",
  "cast_longitude",
  "haul_latitude",
  "haul_longitude",
  "depth_cast",
  "depth_haul",
  "ship_number",
  "wind_speed",
  "wind_direction",
  "weather",
  "cloud",
  "sea_state",
  "bottom_temp",
  "air_temp",
  "barometer",
  "trawl_speed",
  "trawl_length",
  "trawl_bearing",
  "trawl_time",
  "catch",
  "tow_station",
  "quadrant"
)]
```

Treat as categorical variables
```{r}
df_prelim$ship_number <- as.factor(df_prelim$ship_number)
df_prelim$weather <- as.factor(df_prelim$weather)
df_prelim$cloud <- as.factor(df_prelim$cloud)
df_prelim$sea_state <- as.factor(df_prelim$sea_state)
df_prelim$quadrant <- as.factor(df_prelim$quadrant)
```

### Examine catch
```{r}
# Raw catch
hist(df_prelim$catch,
     breaks = 50,
     main = "Distribution of Raw Catch",
     xlab = "Catch (kg)")

# Log-transformed catch
df_prelim$log_catch <- log(df_prelim$catch)

hist(df_prelim$log_catch,
     breaks = 50,
     main = "Distribution of Log(Catch)",
     xlab = "Log(Catch)")
```

### Examine trawl variables
```{r}
#Correlation matrix
trawl_cor <- cor(
  df_prelim[, c(
    "trawl_speed",
    "trawl_length",
    "trawl_time"
  )],
  use = "pairwise.complete.obs"
)

round(trawl_cor, 3)
```
There is essentially no linear correlation between trawl_speed and trawl_length (r=0.055)

There is a moderate negative relationship between trawl_speed and trawl_time (r=-0.483)

There is a strong positive relationship between trawl_length and trawl_time (r=0.768)


Look into relationship between length & time
```{r}
plot(
  df_prelim$trawl_length,
  df_prelim$trawl_time,
  main = "Trawl Length vs Trawl Time",
  xlab = "Trawl Length",
  ylab = "Trawl Time",
  pch = 19
)

abline(
  lm(trawl_time ~ trawl_length,
     data = df_prelim),
  lwd = 2
)
```


### Look into correlations between numerical variables
```{r}
numeric_vars <- df_prelim[, c(
  "year",
  "cast_latitude",
  "cast_longitude",
  "haul_latitude",
  "haul_longitude",
  "depth_cast",
  "depth_haul",
  "wind_speed",
  "wind_direction",
  "bottom_temp",
  "air_temp",
  "barometer",
  "trawl_speed",
  "trawl_length",
  "trawl_bearing",
  "trawl_time"
)]

cor_matrix <- cor(
  numeric_vars,
  use = "pairwise.complete.obs"
)

round(cor_matrix, 2)

corrplot(
  cor_matrix,
  method = "circle",
  type = "upper",
  tl.col = "black",
  tl.srt = 45
)
```

```{r}
preliminary_model <- lm(
  log_catch ~
    year +
    cast_latitude +
    cast_longitude +
    haul_latitude +
    haul_longitude +
    depth_cast +
    depth_haul +
    ship_number +
    wind_speed +
    wind_direction +
    weather +
    cloud +
    sea_state +
    bottom_temp +
    air_temp +
    barometer +
    trawl_speed +
    trawl_length +
    trawl_bearing +
    trawl_time +
    quadrant,
  data = df_prelim
)

summary(preliminary_model)
```

### Check multi collinearity (VIF)
car library not working, make own vif function:
```{r}
# Calculate VIF for each numerical predictor
calculate_vif <- function(model) {
  
  # Get the model matrix
  X <- model.matrix(model)[, -1]
  
  # Calculate VIF for each column
  vif_values <- numeric(ncol(X))
  names(vif_values) <- colnames(X)
  
  for (i in seq_along(vif_values)) {
    
    other_vars <- X[, -i, drop = FALSE]
    target <- X[, i]
    
    auxiliary_model <- lm(target ~ other_vars)
    
    R2 <- summary(auxiliary_model)$r.squared
    
    vif_values[i] <- 1 / (1 - R2)
  }
  
  return(vif_values)
}
```

```{r}
vif_values <- calculate_vif(preliminary_model)
round(vif_values, 2)
```

Rough interpretation of VIF
VIF 1-2, little concern
VIF 2-5, moderate
VIF 5-10, high
VIF >10 Very high

Can remove cast_latitude, cast_longitude as they are highly correlated with haul_latitude, haul_longitude.

Can remove depth_cast as similar to depth_haul 

Even though fround before that trawl speed, time and length are their own variables, the VIF finds speed and time redundant relative to the other trawl variables. Therefore we will remove trawl_speed and trawl_time

Remove quadrant too as it was created from the latitude and longitude variables. This variable may be useful for more simpler models, but not for this multilinear regression model. 

```{r}
# df_prelim$cast_latitude <- NULL
# df_prelim$cast_latitude <- NULL
# df_prelim$depth_cast <- NULL
```

For sake of simplification, we will remove weather, cloud, sea_state, wind_speed, and wind_direction. However, with more time, would try to investigate these variables further. 


## Build second model

### Build new dataset
```{r}
df_model_2 <- df_prelim[, c(
  "year",
  "haul_latitude",
  "haul_longitude",
  "depth_haul",
  "ship_number",
  "bottom_temp",
  "air_temp",
  "barometer",
  "trawl_length",
  "trawl_bearing",
  "catch",
  "log_catch",
  "tow_station"
)]
```

### Build new model
```{r}
model_2 <- lm(
  log_catch ~
    year +
    haul_latitude +
    haul_longitude +
    depth_haul +
    ship_number +
    bottom_temp +
    air_temp +
    barometer +
    trawl_length +
    trawl_bearing,
  data = df_model_2
)

summary(model_2)
```
Model 2 explains 18.36% of variation, but model 1 explains 20.9%.


### Check collinearity
```{r}
vif_values_2 <- calculate_vif(model_2)

round(vif_values_2, 2)
```
Largest VIF is 4.02, which is generally acceptable. 

We can now use model 2 as a baseline model. 

```{r}
summary(model_2)$coefficients
```
barometer is not significant with p=0.897
Similarly, depth_haul also has a high p-value of p=0.964. However this may suggest that the variable is non-linearly correlated with catch.

```{r}
par(mfrow = c(2, 2))
plot(model_2)
```

```{r}
drop1(model_2, test = "F")
```


