---
title: "Dirichlet Model"
author: "Jessica Zeng"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Install once if necessary
install.packages("readxl")
install.packages("dplyr")
install.packages("tidyr")
install.packages("sirt")
```


```{r}
# Load packages
library(readxl)
library(dplyr)
library(tidyr)
library(sirt)
library(ggplot2)
```


```{r}
# File location
file_path <- "./data/spring_historical_english.xlsx"

# Read data
df <- read_excel(file_path)

# Look at the data
head(df)
names(df)
```

Calculate total catch for each year
```{r}
year_total_catch <- df %>%
  group_by(year) %>%
  summarise(
    total_catch = sum(catch, na.rm = TRUE),
    .groups = "drop"
  )

head(year_total_catch)
```

Join total catch back to OG data
```{r}
df_proportions <- df %>%
  left_join(year_total_catch, by = "year")

head(df_proportions)
```

Calculate each stations proportion of yearly catch
```{r}
df_proportions <- df_proportions %>%
  mutate(
    proportion = catch / total_catch
  )


head(df_proportions)
```

Convert to year by station matrix
```{r}
proportion_table <- df_proportions %>%
  select(year, tow_station, proportion) %>%
  pivot_wider(
    names_from = tow_station,
    values_from = proportion,
    values_fill = 0
  ) %>%
  arrange(year)

head(proportion_table)
```

Ensure each year sums to 1
```{r}
row_sums <- proportion_table %>%
  select(-year) %>%
  rowSums()

summary(row_sums)
all(abs(row_sums - 1) < 1e-10)
```

Create matrix for Dirichlet Model
```{r}
proportion_matrix <- proportion_table %>%
  select(-year) %>%
  as.matrix()

dim(proportion_matrix)
```
38 518 means we have 35 years and 518 stations

## Fit Dirichlet Model
```{r}
dirichlet_fit <- dirichlet.mle(proportion_matrix)
```

Get alpha values and put them in a dataframe
```{r}
alpha <- dirichlet_fit$alpha

alpha_table <- data.frame(
  tow_station = colnames(proportion_matrix),
  alpha = as.numeric(alpha)
)

head(alpha_table)
```


Get expected proportion for each station
E[p_i] = alpha_i / sum of all alpha
```{r}
expected_proportions <- dirichlet_fit$xsi

expected_table <- data.frame(
  tow_station = colnames(proportion_matrix),
  expected_proportion = as.numeric(expected_proportions)
)

head(expected_table)
```

Comparison:
Historical average vs fitted Dirichlet expectation
```{r}
historical_means <- colMeans(proportion_matrix)

comparison <- data.frame(
  tow_station = colnames(proportion_matrix),
  historical_mean = historical_means,
  dirichlet_mean = as.numeric(dirichlet_fit$xsi),
  alpha = as.numeric(dirichlet_fit$alpha)
)

head(comparison)
```
They should look similar but will not necessarily be the same

Check concentration param
```{r}
alpha0 <- dirichlet_fit$alpha0

alpha0
```

Tells us about the amount of variation the fitted Dirichlet has around its expected proportions.

Alpha a0 = 754.0381
This means historical proportions tend to stay relatively close to their expected proportions, rather than varying extremely widely. 

Formula: 
alpha_0 = sum of all alphas
E[p_i] = alpha_i / alpha_0



## Generate simulated years
Once the model is fitted, we can generate station-proportion vectos
```{r}
#Adjust this accordingly. Here for 10000 years
n_sim <- 10000

alpha_matrix <- matrix(
  alpha,
  nrow = n_sim,
  ncol = length(alpha),
  byrow = TRUE
)

simulated_proportions <- dirichlet.simul(alpha_matrix)
```

```{r}
dim(simulated_proportions)
```
Each row will be a possible future year.

```{r}
simulated_proportions <- as.data.frame(simulated_proportions)

colnames(simulated_proportions) <- colnames(proportion_matrix)

head(simulated_proportions)
```

Add simulation ID
```{r}
simulated_proportions <- simulated_proportions %>%
  mutate(simulation = row_number()) %>%
  select(simulation, everything())

head(simulated_proportions)

```

Individual station checks
```{r}
station <- "31501"

historical_min <- min(proportion_matrix[, station], na.rm = TRUE)
historical_max <- max(proportion_matrix[, station], na.rm = TRUE)

simulated_min <- min(simulated_proportions[[station]], na.rm = TRUE)
simulated_max <- max(simulated_proportions[[station]], na.rm = TRUE)

c(
  historical_min = historical_min,
  historical_max = historical_max,
  simulated_min = simulated_min,
  simulated_max = simulated_max
)
```


Write to new file
```{r}
write.csv(
  simulated_proportions,
  "simulated_dirichlet_proportions.csv",
  row.names = FALSE
)
```

## Visualisations

### Visualisation 1: One station histroical vs 10000 simulations
```{r}
# Choose a station
station <- "31501"

# Historical proportions
historical_station <- data.frame(
  proportion = proportion_matrix[, station]
)

# Simulated proportions
simulated_station <- data.frame(
  proportion = simulated_proportions[[station]]
)

# Historical min/max
historical_min <- min(historical_station$proportion, na.rm = TRUE)
historical_max <- max(historical_station$proportion, na.rm = TRUE)

ggplot(simulated_station, aes(x = proportion)) +
  
  # Dirichlet simulations
  geom_histogram(
    aes(fill = "Dirichlet simulations"),
    bins = 50,
    boundary = 0,
    alpha = 0.7
  ) +
  
  # Historical minimum
  geom_vline(
    aes(
      xintercept = historical_min,
      linetype = "Historical minimum"
    ),
    colour = "black"
  ) +
  
  # Historical maximum
  geom_vline(
    aes(
      xintercept = historical_max,
      linetype = "Historical maximum"
    ),
    colour = "black"
  ) +
  
  # Historical observations
  geom_rug(
    data = historical_station,
    aes(
      x = proportion,
      colour = "Historical observations"
    ),
    inherit.aes = FALSE,
    sides = "b",
    linewidth = 1
  ) +
  
  scale_fill_manual(
    name = "Data",
    values = c("Dirichlet simulations" = "steelblue")
  ) +
  
  scale_colour_manual(
    name = "Data",
    values = c("Historical observations" = "firebrick")
  ) +
  
  scale_linetype_manual(
    name = "Historical range",
    values = c(
      "Historical minimum" = "dashed",
      "Historical maximum" = "dashed"
    )
  ) +
  
  labs(
    title = paste("Historical vs Dirichlet simulations for station:", station),
    x = "Station proportion of total catch",
    y = "Number of simulations"
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```
Another station
```{r}
# Choose a station
station <- "47404"

# Historical proportions
historical_station <- data.frame(
  proportion = proportion_matrix[, station]
)

# Simulated proportions
simulated_station <- data.frame(
  proportion = simulated_proportions[[station]]
)

# Historical min/max
historical_min <- min(historical_station$proportion, na.rm = TRUE)
historical_max <- max(historical_station$proportion, na.rm = TRUE)

ggplot(simulated_station, aes(x = proportion)) +
  
  # Dirichlet simulations
  geom_histogram(
    aes(fill = "Dirichlet simulations"),
    bins = 50,
    boundary = 0,
    alpha = 0.7
  ) +
  
  # Historical minimum
  geom_vline(
    aes(
      xintercept = historical_min,
      linetype = "Historical minimum"
    ),
    colour = "black"
  ) +
  
  # Historical maximum
  geom_vline(
    aes(
      xintercept = historical_max,
      linetype = "Historical maximum"
    ),
    colour = "black"
  ) +
  
  # Historical observations
  geom_rug(
    data = historical_station,
    aes(
      x = proportion,
      colour = "Historical observations"
    ),
    inherit.aes = FALSE,
    sides = "b",
    linewidth = 1
  ) +
  
  scale_fill_manual(
    name = "Data",
    values = c("Dirichlet simulations" = "steelblue")
  ) +
  
  scale_colour_manual(
    name = "Data",
    values = c("Historical observations" = "firebrick")
  ) +
  
  scale_linetype_manual(
    name = "Historical range",
    values = c(
      "Historical minimum" = "dashed",
      "Historical maximum" = "dashed"
    )
  ) +
  
  labs(
    title = paste("Historical vs Dirichlet simulations for station:", station),
    x = "Station proportion of total catch",
    y = "Number of simulations"
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Compare historical & simulated distributions across all stations
```{r}
historical_summary <- data.frame(
  tow_station = colnames(proportion_matrix),
  mean = colMeans(proportion_matrix),
  sd = apply(proportion_matrix, 2, sd),
  min = apply(proportion_matrix, 2, min),
  max = apply(proportion_matrix, 2, max)
)

simulated_summary <- data.frame(
  tow_station = colnames(simulated_proportions),
  mean = sapply(simulated_proportions, mean),
  sd = sapply(simulated_proportions, sd),
  min = sapply(simulated_proportions, min),
  max = sapply(simulated_proportions, max)
)

comparison_summary <- historical_summary %>%
  rename(
    historical_mean = mean,
    historical_sd = sd,
    historical_min = min,
    historical_max = max
  ) %>%
  left_join(
    simulated_summary %>%
      rename(
        simulated_mean = mean,
        simulated_sd = sd,
        simulated_min = min,
        simulated_max = max
      ),
    by = "tow_station"
  )
```

Historical mean vs Dirichlet simulated mean
```{r}
ggplot(
  comparison_summary,
  aes(x = historical_mean, y = simulated_mean)
) +
  geom_point(
    aes(colour = "Station"),
    alpha = 0.7,
    size = 2
  ) +
 geom_abline(
  slope = 1,
  intercept = 0,
  aes(linetype = "Perfect agreement"),
  colour = "black"
  ) +
  scale_colour_manual(
    name = "Data",
    values = c("Station" = "steelblue")
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("Perfect agreement" = "dashed")
  ) +
  labs(
    title = "Historical vs Dirichlet Mean Proportions",
    x = "Historical mean proportion",
    y = "Dirichlet simulated mean proportion"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```


```{r}
ggplot(
  comparison_summary,
  aes(x = historical_sd, y = simulated_sd)
) +
  geom_point(
    aes(colour = "Station"),
    alpha = 0.7,
    size = 2
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    aes(linetype = "Perfect agreement"),
    colour = "black"
  ) +
  scale_colour_manual(
    name = "Data",
    values = c("Station" = "steelblue")
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("Perfect agreement" = "dashed")
  ) +
  labs(
    title = "Historical vs Dirichlet Variability",
    x = "Historical standard deviation",
    y = "Dirichlet simulated standard deviation"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Historical minimum vs Dirichlet simulated minimum
```{r}
ggplot(
  comparison_summary,
  aes(x = historical_min, y = simulated_min)
) +
  geom_point(
    aes(colour = "Station"),
    alpha = 0.7,
    size = 2
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    aes(linetype = "Perfect agreement"),
    colour = "black"
  ) +
  scale_colour_manual(
    name = "Data",
    values = c("Station" = "steelblue")
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("Perfect agreement" = "dashed")
  ) +
  labs(
    title = "Historical vs Dirichlet Minimum Proportions",
    x = "Historical minimum proportion",
    y = "Dirichlet simulated minimum proportion"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```


Historical maximum vs Dirichlet simulated maximum

```{r}
ggplot(
  comparison_summary,
  aes(x = historical_max, y = simulated_max)
) +
  geom_point(
    aes(colour = "Station"),
    alpha = 0.7,
    size = 2
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    aes(linetype = "Perfect agreement"),
    colour = "black"
  ) +
  scale_colour_manual(
    name = "Data",
    values = c("Station" = "steelblue")
  ) +
  scale_linetype_manual(
    name = "Reference",
    values = c("Perfect agreement" = "dashed")
  ) +
  labs(
    title = "Historical vs Dirichlet Maximum Proportions",
    x = "Historical maximum proportion",
    y = "Dirichlet simulated maximum proportion"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

## VERIFYING
Use this model to predict catch in one year.
```{r}


```

