---
title: "Station by Station Historical Model"
author: "Jessica Zeng"
date: "2026-09-10"
output: html_document
---

```{r}
# Install once if necessary
install.packages("readxl")
install.packages("dplyr")
install.packages("tidyr")
install.packages("sirt")
install.packages("MCMCpack")
```


```{r}
# Load packages
library(readxl)
library(dplyr)
library(tidyr)
library(sirt)
library(ggplot2)
library(MCMCpack)
```

```{r}
set.seed(566)
```


Reading in Data
```{r}
file_path <- "./data/spring_historical_english.xlsx"
df <- read_excel(file_path)
head(df)
names(df)
```

Extract only columns we want
```{r}
# df %>%
#   dplyr::select(year, tow_station, catch) %>%
#   head(20)
df <- df[, c("year", "tow_station", "catch")]
```

Check for missing values
```{r}
df %>%
  summarise(
    missing_year = sum(is.na(year)),
    missing_station = sum(is.na(tow_station)),
    missing_catch = sum(is.na(catch))
  )
```

Plot total catch
```{r}
annual_catch <- df %>%
  group_by(year) %>%
  summarise(
    total_catch = sum(catch, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(annual_catch, aes(x = year, y = total_catch)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Total Catch Across All Stations by Year",
    x = "Year",
    y = "Total Catch (kg)"
  ) +
  theme_minimal()
```
Truncate starting from year 2002
```{r}
truncation_year <- 2002

df_truncated <- df %>%
  filter(year <= truncation_year)
```


Station by year catch matrix
```{r}
catch_matrix <- df[, c("year", "tow_station", "catch")] %>%
  pivot_wider(
    names_from = tow_station,
    values_from = catch
  )

head(catch_matrix)
```

Dirichlet function - rdirichlet
- A Dirichlet random vector can be generated using independent Gamma random variables X_i = Gamma (alpha_i, 1)
- The function allows us to generate random numbers that sum up to 1. 
n = how many dirichlet random vectors we want to generate. i.e sets of proportions
alpha = (1,1,...,1,1) means all elements in the array have the same concentration parameter. 


## Apply to ONE station
```{r}
example_station <- df %>%
  filter(tow_station == unique(df$tow_station)[3]) %>%
  .[, c("year", "tow_station", "catch")] %>%
  filter(!is.na(catch)) %>%
  arrange(year)

example_station
```


Extract historical catch
```{r}
historical_catch <- example_station$catch
historical_catch
```
Generate Dirichlet distributions
```{r}
alpha <- rep(1, length(historical_catch))


proportions <- rdirichlet(
  n = 1,
  alpha = alpha
)

proportions
```
Explanation: these proportions are the proportions across multiple years catch for ONE station. 

```{r}
simulated_catch <- sum(
  historical_catch * proportions
)

simulated_catch
```

Generate many simlations for one station
```{r}
n_simulations <- 100

proportions <- rdirichlet(
  n = n_simulations,
  alpha = rep(1, length(historical_catch))
)
```

Proportions is now a matrix
- Each row: A simulation 
- Each column: Each year

```{r}
simulated_catches <- proportions %*% historical_catch

simulated_catches <- as.vector(simulated_catches)

summary(simulated_catches)
```


Compare with historical catch for the same station:
```{r}
summary(historical_catch)
```
```{r}
simulation_df <- data.frame(
  catch = simulated_catches
)

ggplot() +
  geom_histogram(
    data = data.frame(catch = historical_catch),
    aes(x = catch, fill = "Historical"),
    bins = 20,
    alpha = 0.5
  ) +
  geom_histogram(
    data = simulation_df,
    aes(x = catch, fill = "Simulated"),
    bins = 20,
    alpha = 0.5
  ) +
  labs(
    title = "Historical vs Simulated Catch",
    x = "Catch",
    y = "Frequency",
    fill = "Data"
  ) +
  scale_fill_manual(
    values = c(
      "Historical" = "steelblue",
      "Simulated" = "orange"
    )
  ) +
  theme_minimal()
```

```{r}
example_station <- df %>%
  filter(tow_station == unique(df$tow_station)[1]) %>%
  .[, c("year", "tow_station", "catch")] %>%
  filter(!is.na(catch)) %>%
  arrange(year)

random_year <- sample(
  truncation_year:2020,
  1
)

historical_value <- example_station %>%
  filter(year == random_year) %>%
  pull(catch)

random_year
historical_value


lower_95 <- quantile(simulated_catches, 0.025)
upper_95 <- quantile(simulated_catches, 0.975)
```


```{r}
simulation_df <- data.frame(
  catch = simulated_catches
)

ggplot(simulation_df, aes(x = catch)) +
  
  # Simulated catch distribution
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    alpha = 0.7
  ) +
  
  # 95% confidence/credible interval
  geom_vline(
    xintercept = lower_95,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  geom_vline(
    xintercept = upper_95,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  
  # Historical catch
  geom_vline(
    xintercept = historical_value,
    linewidth = 1.2
  ) +
  
  labs(
    title = paste(
      "Simulated Catch vs Historical Catch — Year",
      random_year
    ),
    subtitle = paste(
      "95% simulation interval:",
      round(lower_95, 1), "–", round(upper_95, 1), "kg"
    ),
    x = "Catch (kg)",
    y = "Frequency"
  ) +
  
  theme_minimal()
```

## Verifying
How often do we fall inside the interval?

```{r}
example_station <- df %>%
  filter(
    tow_station == unique(df$tow_station)[1],
    !is.na(catch)
  ) %>%
  arrange(year)

# Data used to build the Dirichlet model
training_catch <- example_station %>%
  filter(year <= truncation_year) %>%
  pull(catch)

# Data used to test the model
validation_data <- example_station %>%
  filter(year > truncation_year)

last_year <- max(example_station$year)
```

```{r}
n_repetitions <- 1000
n_simulations <- 10000
alpha_value <- 1

validation_results <- data.frame(
  repetition = integer(n_repetitions),
  year = integer(n_repetitions),
  historical_catch = numeric(n_repetitions),
  lower_95 = numeric(n_repetitions),
  upper_95 = numeric(n_repetitions),
  inside_interval = logical(n_repetitions)
)

for (i in 1:n_repetitions) {
  
  # Randomly select one validation year
  selected_year <- sample(validation_data$year, 1)
  
  # Get the historical catch for that year
  historical_value <- validation_data %>%
    filter(year == selected_year) %>%
    pull(catch)
  
  # Equal alpha for every historical year
  alpha <- rep(alpha_value, length(training_catch))
  
  # Generate Dirichlet proportions
  proportions <- rdirichlet(
    n = n_simulations,
    alpha = alpha
  )
  
  # Convert proportions into simulated catches
  simulated_catches <- as.vector(
    proportions %*% training_catch
  )
  
  # Calculate 95% simulation interval
  lower_95 <- quantile(simulated_catches, 0.025)
  upper_95 <- quantile(simulated_catches, 0.975)
  
  # Check whether historical catch is inside interval
  inside <- (
    historical_value >= lower_95 &&
    historical_value <= upper_95
  )
  
  # Store results
  validation_results[i, ] <- c(
    i,
    selected_year,
    historical_value,
    lower_95,
    upper_95,
    inside
  )
}
```

```{r}
validation_summary <- validation_results %>%
  summarise(
    inside = sum(inside_interval),
    outside = sum(!inside_interval),
    total = n(),
    percentage_inside = mean(inside_interval) * 100,
    percentage_outside = mean(!inside_interval) * 100
  )

validation_summary
```

# All stations
```{r warning=FALSE}
validation_results <- list()

station_list <- unique(df$tow_station)

for (station in station_list) {
  
  # Get data for this station
  station_data <- df %>%
    filter(
      tow_station == station,
      !is.na(catch)
    ) %>%
    arrange(year)
  
  # Historical data used to build the model
  training_catch <- station_data %>%
    filter(year <= truncation_year) %>%
    pull(catch)
  
  # Historical data used for validation
  validation_data <- station_data %>%
    filter(year > truncation_year)
  
  # Skip station if there is insufficient data
  if (length(training_catch) < 2 || nrow(validation_data) == 0) {
    next
  }
  
  # --------------------------------------------------
  # Generate simulated catch distribution
  # --------------------------------------------------
  
  alpha <- rep(alpha_value, length(training_catch))
  
  proportions <- rdirichlet(
    n = n_simulations,
    alpha = alpha
  )
  
  simulated_catches <- as.vector(
    proportions %*% training_catch
  )
  
  # 95% simulation interval
  lower_95 <- quantile(simulated_catches, 0.025)
  upper_95 <- quantile(simulated_catches, 0.975)
  
  # --------------------------------------------------
  # Repeatedly select historical validation years
  # --------------------------------------------------
  
  selected_years <- sample(
    validation_data$year,
    size = n_repetitions,
    replace = TRUE
  )
  
  # Get historical catches corresponding to selected years
  historical_values <- validation_data$catch[
    match(selected_years, validation_data$year)
  ]
  
  # Determine whether each historical catch
  # falls inside the simulated 95% interval
  inside <- (
    historical_values >= lower_95 &
    historical_values <= upper_95
  )
  
  # Store results
  validation_results[[as.character(station)]] <- data.frame(
    tow_station = station,
    repetition = 1:n_repetitions,
    year = selected_years,
    historical_catch = historical_values,
    lower_95 = lower_95,
    upper_95 = upper_95,
    inside_interval = inside
  )
}

# Combine all stations
validation_results <- bind_rows(validation_results)
```


```{r}
station_summary <- validation_results %>%
  group_by(tow_station) %>%
  summarise(
    inside = sum(inside_interval),
    outside = sum(!inside_interval),
    total = n(),
    percentage_inside = mean(inside_interval) * 100,
    percentage_outside = mean(!inside_interval) * 100,
    .groups = "drop"
  )

station_summary
```


```{r}
ggplot(
  station_summary,
  aes(x = percentage_inside)
) +
  geom_histogram(
    bins = 20
  ) +
  labs(
    title = "95% Simulation Interval Coverage Across Stations",
    x = "Historical catches inside 95% interval (%)",
    y = "Number of stations"
  ) +
  theme_minimal()
```

